---
title: "HW3: Pairwise Alignment Solutions" 
author: "Author: Your Name"
date: "Last update: April 23, 2026" 
output:
  html_document:
    toc: true
    toc_float:
        collapsed: true
        smooth_scroll: true
    toc_depth: 3
    fig_caption: yes
    code_folding: show
    number_sections: false
fontsize: 14pt
bibliography: bibtex.bib
type: docs
weight: 304
---

<!---
- Compile from command-line
Rscript -e "rmarkdown::render('HW3_key.Rmd', c('html_document', 'pdf_document'), clean=FALSE)"
-->

<div style="text-align: right"> 
Source code downloads: &nbsp; &nbsp;
[ [.Rmd](https://raw.githubusercontent.com/tgirke/GEN242/refs/heads/main/assignments/homework/hw03/hw3_solution/HW3_key.Rmd) ] &nbsp; &nbsp; 
[ [pairwiseAlign_Fct.R](https://raw.githubusercontent.com/tgirke/GEN242/refs/heads/main/assignments/homework/hw03/hw3_solution/pairwiseAlign_Fct.R) ]
</div>

# Rendering Instructions 

To render this R Markdown document, one needs to download the following files to the same directory.

   + [HW3_key.Rmd](https://girke.bioinformatics.ucr.edu/GEN242/assignments/homework/hw03/hw3_solution/HW3_key.html): Rmd source file for this document
   + [pairwiseAlign_Fct.R](): R script defining pairwise alignment functions
   + [bibtex.bib](https://raw.githubusercontent.com/tgirke/GEN242/refs/heads/main/assignments/homework/hw03/hw3_solution/bibtex.bib): references cited in the text in BibTeX format

Next, one can render the report to HTML, PDF and other formats following the instructions below. Both the HTML and PDF versions are linked. Note, the rendering needs to be performed from the same directory where the downloaded `*.Rmd`, `*.R` and `*.bib` files are located.

   + [HTML](https://girke.bioinformatics.ucr.edu/GEN242/assignments/homework/hw03/hw3_solution/HW3_key.html): this report in HTML format
   + [PDF](https://girke.bioinformatics.ucr.edu/GEN242/assignments/homework/hw03/hw3_solution/HW3_key.pdf): corresponding PDF version


The HTML report can be rendered with `rmarkdown::render()` as follows. 

```{r render_this_markdown, eval=FALSE, message=FALSE}
rmarkdown::render('HW4_key.Rmd') # From R
Roscript -e "rmarkdown::render('HW4_key.Rmd')" # From command-line
```

To render a PDF file instead of HTML, one can instruct the rendering function to do so like this: `rmarkdown::render('HW4_key.Rmd', c('pdf_document')`. To render to several formats with 
a single command, one can concatenate the formatting values with `c('html_document', 'pdf_document')`.  

# A. Choice of Sequence Type
Task 1: Which sequence type - amino acid or nucleotide - is more appropriate to search databases 
for remotely related sequences? Provide at least three reasons for your decision.

Answer: When coding sequences are expected to have weak similarities then one
should use protein sequences rather than DNA sequences for database searching, 
because of (1) their higher information content (20 versus 4 letter alphabet), 
as well as (2) the better scoring and (3) functional classification systems available 
for amino acids.

# B. Dynamic Programming for Pairwise Alignments

Task 2: Create manually (or write an R script for it) one global and one local
alignment for the following two protein sequences using the Needleman-Wusch and
Smith-Waterman algorithms, respectively [@Smith1981-ax; @Needleman1970-md].

```sh
O15528: PFGFGKRSCMGRRLA
P98187: FIPFSAGPRNCIGQK
```

## Source functions 

All alignment functions used in the following sections are defined in the downloaded R script file
that is named `pairwiseAlign_Fct.R`. These functions are loaded with the `source()` command below.

```{r source_pairalign_fct, eval=TRUE, message=FALSE}
source("pairwiseAlign_Fct.R")
```

## Input sequences

Define within R or import them (here former).

```{r input_seq1, eval=TRUE}
S1 <- "PFGFGKRSCMGRRLA"
S2 <- "FIPFSAGPRNCIGQK"
```

Additional test sequences

```{r input_seq2, eval=FALSE}
# S1 <- "HEAGAWGHEE"
# S2 <- "PAWHEAE"
```

## Global alignment 

The alignment type choice is passed on to all following functions.

```{r align_type, eval=TRUE, message=FALSE}
align_type <- "global"
# align_type <- "local"
```

### Dynamic programming matrices 

```{r gen_dynProgMatrix, eval=TRUE, message=FALSE}
dynMA <- dynProgMatrix(S1, S2, align_method=align_type, gap_penalty=8, substitutionMA="BLOSUM50")
```

The matrices are stored in a list and returned below. The path is indicated by three numbers in the `glob_ma_path` matrix.
Their meaning is: 

+ 1: diagonal
+ 2: vertical (up)
+ 3: horizontal (left)

```{r print_dynProgMatrix, eval=TRUE, message=FALSE}
dynMA
```

### Compute alignment

The following `alignList` stores all relevant results in a list, including dynamic 
programming matrices, as well as the coordinates (named `path_coor`) to highlight path in 
dynamic progamming matrix (see below). 

```{r gen_alignmentTraceback, eval=TRUE, message=FALSE}
alignList <- alignmentTraceback(ma=dynMA[[1]], ma_path=dynMA[[2]], align_method=align_type) 
names(alignList)
# alignList$ma # dyn ma with scores
# alignList$ma_path # dyn ma with path
# alignList$path_coor # coordinates for path to auto highlight path in HTML/PDF table
```

### Return results

#### Traceback in matrix 

The following prints the fully populated dynamic programming matrix where the traceback 
path is highlighted in color.

```{r gen_alignmentTraceback_Matrix, eval=TRUE, message=FALSE}
printColMa(alignList)
```

#### Alignment and score

```{r gen_alignment, eval=TRUE, message=FALSE}
printAlign(x=alignList)
```

## Local alignment 

The alignment type choice is passed on to all following functions.

```{r align_type_loc, eval=TRUE, message=FALSE}
# align_type <- "global"
align_type <- "local"
```

### Dynamic programming matrices 

```{r gen_dynProgMatrix_loc, eval=TRUE, message=FALSE}
dynMA <- dynProgMatrix(S1, S2, align_method=align_type, gap_penalty=8, substitutionMA="BLOSUM50")
```

The matrices are stored in a list and returned below.

```{r print_dynProgMatrix_loc, eval=TRUE, message=FALSE}
dynMA
```

### Compute alignment

Note: `alignList` stores all relevant results in a list, including dynamic 
programming matrices, as well as the coordinates (named `path_coor`) to highlight the path in 
the dynamic progamming matrix. This way one can easily generate a single dynamic programming matrix
with the traceback path highlighted by colors or arrows in an HTML or PDF document (see below).

```{r gen_alignmentTraceback_loc, eval=TRUE, message=FALSE}
alignList <- alignmentTraceback(ma=dynMA[[1]], ma_path=dynMA[[2]], align_method=align_type) 
names(alignList)
# alignList$ma # dyn ma with scores
# alignList$ma_path # dyn ma with path
# alignList$path_coor # coordinates for path to auto highlight path in HTML/PDF table
```

### Return results

#### Traceback in matrix 

The following prints the fully populated dynamic programming matrix where the traceback 
path is highlighted in color.

```{r gen_alignmentTraceback_Matrix_loc, eval=TRUE, message=FALSE}
printColMa(alignList)
```

#### Alignment and score

```{r gen_alignment_loc, eval=TRUE, message=FALSE}
printAlign(x=alignList)
```

# C. Different Substitution Matrices
Task 1: Load the Biostrings/pwalign packages in R, import the following two cytochrome
P450 sequences from NCBI: O15528 and P98187 (save as myseq.fasta), and create a
global alignment with the `pairwiseAlignment` function from pwalign.

Instead of downloading the sequences manually from NCBI, one can use the rentrez package 
to load the sequences directly into R as follows. Note, this will save the sequences to 
a FASTA file (named myseq.fasta) in your current directory, that will be loaded in the 
next step to a `AAStringSet` object.

```{r rentrez_download, eval=FALSE, message=FALSE}
library(rentrez)
myseq <- entrez_fetch(db="protein", id=c("O15528", "P98187"), rettype="fasta")
write(myseq, file="myseq.fasta") # Stores sequences in myseq.fasta
```

The following creates the pairwise alignments. 

```{r alignment_scores_biostrings, eval=FALSE, message=FALSE}
library(Biostrings); library(pwalign)
myseq <- readAAStringSet("myseq.fasta", "fasta")
(p <- pairwiseAlignment(myseq[[1]], myseq[[2]], type="global", substitutionMatrix="BLOSUM50"))
writePairwiseAlignments(p)
```

Your answers should address the following tasks:

Record the scores for the scoring matrices BLOSUM50, BLOSUM62 and BLOSUM80.
How and why do the scores differ for the three scoring matrices?

Answer 1: The scores for the three BLOSUM substitutions matrices are:

+ BLOSUM50: 227 
+ BLOSUM62: 54
+ BLOSUM80: -52

Answer 2: Since the two sequences are relatively dissimilar (as determined by alignment view from
`writePairwiseAlignments(p)`) it is expected that the BLOSUM matrices trained on less similar
sequences (_e.g._ BLOSUM50) result in higher scores than those trained on more similar 
sequences (_e.g._ BLOSUM80).


# Session Info

```{r sessionInfo}
sessionInfo()
```

# References

