---
title: "NGS Analysis Basics"
subtitle: "GEN242: Data Analysis in Genome Biology"
author: "Thomas Girke"
date: today
format:
  revealjs:
    theme: [default, ../../assets/revealjs_custom.scss]
    slide-number: true
    progress: true
    scrollable: true
    smaller: true
    highlight-style: github
    code-block-height: 380px
    transition: slide
    embed-resources: true
    footer: "GEN242 · UC Riverside · [Tutorial](https://girke.bioinformatics.ucr.edu/GEN242/tutorials/rsequences/rsequences_index.html) · [HW05](https://girke.bioinformatics.ucr.edu/GEN242/assignments/homework/hw05/hw05/)"
    logo: "https://girke.bioinformatics.ucr.edu/GEN242/assets/logo_gen242.png"
execute:
  echo: true
  eval: true
---

## Overview

Topics covered in this tutorial:

- String handling in R base
- Key Bioconductor packages for sequence analysis
- Biostrings: sequence containers, import/export, manipulation
- Pattern matching with and without mismatches
- Sequence logos and PWM searching
- NGS data: FASTQ format and quality scores
- FASTQ quality reports and QC
- Filtering and trimming FASTQ files
- Memory-efficient FASTQ processing
- Range operations with `IRanges` and `GRanges`
- Transcript ranges and `TxDb` databases
- Efficient sequence parsing from genome files

::: {.callout-note}
**HW06** tasks are linked at the end of this tutorial.  
Full assignment: [HW06](https://girke.bioinformatics.ucr.edu/GEN242/assignments/homework/hw06/hw06/)
:::

---

## Key Bioconductor Packages for Sequence Analysis

### R Base

Basic string handling (`grep`, `gsub`, `substring`) plus the full range of
numeric data analysis tools.

### Bioconductor — sequence-specific infrastructure

| Package | Purpose |
|---|---|
| [Biostrings](http://bioconductor.org/packages/release/bioc/html/Biostrings.html) | General sequence analysis — the core package |
| [ShortRead](http://bioconductor.org/packages/release/bioc/html/ShortRead.html) | Pipeline for short read / FASTQ data |
| [IRanges](http://bioconductor.org/packages/release/bioc/html/IRanges.html) | Low-level range data infrastructure |
| [GenomicRanges](http://bioconductor.org/packages/release/bioc/html/GenomicRanges.html) | High-level genomic range data |
| [GenomicFeatures](http://bioconductor.org/packages/release/bioc/html/GenomicFeatures.html) | Transcript-centric annotation management |
| [GenomicAlignments](http://bioconductor.org/packages/release/bioc/html/GenomicAlignments.html) | Short genomic alignments |
| [Rsamtools](http://bioconductor.org/packages/release/bioc/html/Rsamtools.html) | Interface to `samtools`, `bcftools`, `tabix` |
| [BSgenome](http://bioconductor.org/packages/release/bioc/html/BSgenome.html) | Genome annotation data |
| [biomaRt](http://bioconductor.org/packages/release/bioc/html/biomaRt.html) | Interface to BioMart annotations |
| [rtracklayer](http://bioconductor.org/packages/release/bioc/html/rtracklayer.html) | Annotation imports, interface to genome browsers |

### Install required packages

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(c("Biostrings", "GenomicRanges", "rtracklayer",
                        "systemPipeR", "seqLogo", "ShortRead"))
```

---

## String Handling in R Base {.scrollable}

### Sample sequence data

```{r}
myseq <- c("ATGCAGACATAGTG", "ATGAACATAGATCC", "GTACAGATCAC")
```

### String matching

```{r}
myseq[grep("ATG", myseq)]                        # sequences containing "ATG"

pos1 <- regexpr("AT", myseq)                     # position of FIRST match per sequence
as.numeric(pos1)
attributes(pos1)$match.length

pos2 <- gregexpr("AT", myseq)                    # positions of ALL matches per sequence
as.numeric(pos2[[1]])                            # match positions in first sequence
attributes(pos2[[1]])$match.length               # match lengths in first sequence
```

### String substitution

```{r}
gsub("^ATG", "atg", myseq)                       # replace "ATG" at start of string
```

### Positional parsing

```{r}
nchar(myseq)                                      # length of each string
substring(myseq[1], c(1,3), c(2,5))              # extract two fragments from one string
substring(myseq, c(1,4,7), c(2,6,10))            # one fragment from each string
```

### Generate random DNA sequences

```{r}
rand <- sapply(1:100, function(x)
    paste(sample(c("A","T","G","C"), sample(10:20, 1), replace=TRUE), collapse=""))
rand[1:3]
table(c(rand[1:4], rand[1]))                      # count identical sequences
```

---

## Biostrings — Data Objects {.scrollable}

The `Biostrings` package provides dedicated containers for biological sequences.

### `XString` — single sequence

| Class | Use |
|---|---|
| `DNAString` | DNA sequence |
| `RNAString` | RNA sequence |
| `AAString` | Amino acid sequence |
| `BString` | Any character string |

```{r}
library(Biostrings)
d <- DNAString("GCATAT-TAC")
d
d[1:4]                        # subsetting by position

r <- RNAString(d)             # convert DNA to RNA
p <- AAString("HCWYHH")       # amino acid sequence
b <- BString("Any characters can go here.")
```

### `XStringSet` — many sequences in one object

| Class | Use |
|---|---|
| `DNAStringSet` | set of DNA sequences |
| `RNAStringSet` | set of RNA sequences |
| `AAStringSet` | set of amino acid sequences |

```{r}
dset <- DNAStringSet(c("GCATATTAC", "AATCGATCC", "GCATATTAC"))
names(dset) <- c("seq1", "seq2", "seq3")
dset[1:2]

width(dset)                    # length of each sequence
d <- dset[[1]]                 # [[ returns single entry as XString
dset2 <- c(dset, dset)         # concatenate two XStringSet objects
dsetchar <- as.character(dset) # convert to named character vector
dsetone <- unlist(dset)        # collapse all sequences into one DNAString

DNAStringSet(dset, start=c(1,2,3), end=c(4,8,5))  # subset by positions
```

### `QualityScaleXStringSet` — sequences with quality scores

```{r}
# QualityScaledDNAStringSet, QualityScaledRNAStringSet, QualityScaledAAStringSet
```

---

## Biostrings — Import, Export and Manipulation {.scrollable}

### Import FASTA file

```{r}
# Download sample sequences
dir.create("data", showWarnings = FALSE)
download.file(
  "https://ftp.ncbi.nlm.nih.gov/genomes/archive/old_genbank/Bacteria/Halobacterium_sp_uid217/AE004437.ffn",
  "data/AE004437.ffn"
)
myseq <- readDNAStringSet("data/AE004437.ffn")  # import FASTA
myseq[1:3]
```

### Subset and export

```{r}
sub <- myseq[grep("99.*", names(myseq))]         # subset by name pattern
length(sub)
writeXStringSet(sub, file="./data/AE004437sub.ffn", width=80)  # export to FASTA
```

### Basic sequence manipulations

```{r}
randset <- DNAStringSet(rand)                     # convert random seqs to DNAStringSet
complement(randset[1:2])                          # complement
reverse(randset[1:2])                             # reverse
reverseComplement(randset[1:2])                   # reverse complement (standard for NGS)
translate(randset[1:2])                           # translate DNA to protein
```

### Multiple sequence alignment

```{r}
#| eval: false
origMAlign <- readDNAMultipleAlignment(
    filepath = system.file("extdata", "msx2_mRNA.aln", package="Biostrings"),
    format = "clustal"
)
origMAlign
```

---

## Pattern Matching with Biostrings {.scrollable}

Biostrings supports pattern matching with **mismatches** — not possible with base R regex.

### Match a single pattern against one sequence

```{r}
myseq1 <- readDNAStringSet("./data/AE004437.ffn")

# Find all occurrences allowing 1 mismatch
mypos <- matchPattern("ATGGTG", myseq1[[1]], max.mismatch=1)

# Count matches only
countPattern("ATGGCT", myseq1[[1]], max.mismatch=1)
```

### Match against all sequences in a set

```{r}
vcountPattern("ATGGCT", myseq1, max.mismatch=1)[1:20]   # count per sequence

myvpos <- vmatchPattern("ATGGCT", myseq1, max.mismatch=1) # match positions (MIndex)
Views(myseq1[[1]], start(myvpos[[1]]), end(myvpos[[1]]))   # retrieve matches for seq 1

# Retrieve match sequences across all entries
sapply(names(myseq1), function(x)
    as.character(Views(myseq1[[x]], start(myvpos[[x]]), end(myvpos[[x]]))))[1:4]
```

### Consensus matrix for query and hits

```{r}
tmp <- c(DNAStringSet("ATGGTG"), DNAStringSet(mypos))
consensusMatrix(tmp)[1:4,]    # shows nucleotide frequency at each position
```

### Pattern matching with regular expression support

```{r}
myseq <- DNAStringSet(c("ATGCAGACATAGTG", "ATGAACATAGATCC", "GTACAGATCAC"))
myseq[grep("^ATG", myseq, perl=TRUE)]            # sequences starting with ATG
DNAStringSet(gsub("^ATG", "NNN", myseq))         # substitute ATG at start with NNN
```

---

## Sequence Logos and PWM Searching

### Build a position weight matrix (PWM)

```{r}
library(seqLogo)
pwm <- PWM(DNAStringSet(c("GCT", "GGT", "GCA")))
pwm
```

### Plot sequence logo with `seqLogo`

```{r}
seqLogo(t(t(pwm) * 1/colSums(pwm)))
```

### Plot with `ggseqlogo` (more customization options)

```{r}
library(ggplot2); library(ggseqlogo)
pwm <- PWM(DNAStringSet(c("GCT", "GGT", "GCA")))
ggseqlogo(pwm)
```

`ggseqlogo` supports various alphabets including amino acid sequences.
See the [manual](https://omarwagih.github.io/ggseqlogo/) for details.

### Search a sequence for PWM matches

```{r}
chr <- DNAString("AAAGCTAAAGGTAAAGCAAAA")
matchPWM(pwm, chr, min.score=0.9)   # returns matches scoring > 90% of maximum
```

---

## FASTQ Format — Sequence and Quality Data {.scrollable}

FASTQ is the standard format for raw NGS reads. Each read occupies **4 lines**:

```default
@SRR038845.3 HWI-EAS038:6:1:0:1938 length=36
CAACGAGTTCACACCTTGGCCGACAGGCCCGGGTAA
+SRR038845.3 HWI-EAS038:6:1:0:1938 length=36
BA@7>B=>:>>7@7@>>9=BAA?;>52;>:9=8.=A
```

| Line | Content |
|---|---|
| 1 | `@` + read ID and description |
| 2 | DNA sequence |
| 3 | `+` (optionally repeated ID) |
| 4 | Quality scores as ASCII characters (Phred encoding) |

### Phred quality scores

Phred scores are integers 0–50 stored as ASCII characters (score + 33):

```{r}
phred <- 1:9
phreda <- paste(sapply(as.raw((phred)+33), rawToChar), collapse="")
phreda                           # ASCII representation
as.integer(charToRaw(phreda))-33 # convert back to integers
```

- **Phred 10** = 90% base call accuracy (1 error in 10)
- **Phred 20** = 99% accuracy (1 error in 100) — common QC cutoff
- **Phred 30** = 99.9% accuracy (1 error in 1,000)

---

## Working with FASTQ Files — ShortRead {.scrollable}

### Download sample data and import

```{r}
library(ShortRead)
download.file(
    "https://cluster.hpcc.ucr.edu/~tgirke/HTML_Presentations/Manuals/testdata/samplefastq/data.zip",
    "data.zip"
)
unzip("data.zip")

fastq <- list.files("data", "*.fastq$")
fastq <- paste("data/", fastq, sep="")
names(fastq) <- paste("flowcell6_lane", 1:length(fastq), sep="_")
```

### Read and inspect a FASTQ file

```{r}
fq <- readFastq(fastq[1])          # import first FASTQ file as ShortReadQ object
countLines(dirPath="./data", pattern=".fastq$") / 4  # count reads in each file

id(fq)[1]                          # read ID
sread(fq)[1]                       # sequence
quality(fq)[1]                     # Phred quality scores (ASCII)
as(quality(fq), "matrix")[1:4, 1:12]  # convert Phred to numeric matrix

ShortReadQ(sread=sread(fq), quality=quality(fq), id=id(fq))  # construct from components
```

### Construct a `QualityScaledDNAStringSet` from scratch

```{r}
dset <- DNAStringSet(sapply(1:100, function(x)
    paste(sample(c("A","T","G","C"), 20, replace=TRUE), collapse="")))
myqlist <- lapply(1:100, function(x) sample(1:40, 20, replace=TRUE))
myqual <- sapply(myqlist, function(x) toString(PhredQuality(x)))
myqual <- PhredQuality(myqual)
dsetq1 <- QualityScaledDNAStringSet(dset, myqual)
dsetq1[1:2]
```

---

## FASTQ Quality Reports {.scrollable}

### Using `systemPipeR` to create comprehensive QC plots

```{r see_fastq}
#| eval: true
#| message: false
#| warning: false
#| fig-width: 22
#| fig-height: 12
library(systemPipeR)
fqlist <- seeFastq(fastq=fastq, batchsize=800, klength=8) # For real data set batchsize to at least 10^5 
seeFastqPlot(fqlist)
```

Generates a multi-panel QC report covering:
- Read length distribution
- Per-position base quality
- Per-position nucleotide frequency
- K-mer frequency
- Adapter contamination

Output can be written to a single PDF covering all samples.

### Using `ShortRead` — alternative QC

```{r}
sp <- SolexaPath(system.file('extdata', package='ShortRead'))
fl <- file.path(analysisPath(sp), "s_1_sequence.txt")
fls <- c(fl, fl)
coll <- QACollate(
    QAFastqSource(fls),
    QAReadQuality(),
    QAAdapterContamination(),
    QANucleotideUse(),
    QAQualityUse(),
    QASequenceUse(),
    QAFrequentSequence(n=10),
    QANucleotideByCycle(),
    QAQualityByCycle()
)
x <- qa2(coll, verbose=TRUE)
res <- report(x)
if (interactive()) browseURL(res)   # open HTML report in browser
```

---

## Filtering and Trimming FASTQ Files {.scrollable}

All filtering operates on `ShortReadQ` objects and returns a filtered subset.

### Adapter trimming

```{r}
fqtrim <- trimLRPatterns(Rpattern="GCCCGGGTAA", subject=fq)
sread(fq)[1:2]      # before trimming
sread(fqtrim)[1:2]  # after trimming — adapter sequence removed
```

### Count reads and remove duplicates

```{r}
tables(fq)$distribution          # frequency table of read occurrences
sum(srduplicated(fq))            # number of duplicated reads
fq[!srduplicated(fq)]           # keep only unique reads
```

### Trim low-quality tails

```{r}
cutoff <- 30
cutoff <- rawToChar(as.raw(cutoff + 33))   # convert Phred score to ASCII
sread(trimTails(fq, k=2, a=cutoff, successive=FALSE))[1:2]
# k=2: trim if 2 consecutive bases fall below cutoff
```

### Remove reads with low average quality

```{r}
cutoff <- 30
qcount <- rowSums(as(quality(fq), "matrix") <= 20)
fq[qcount == 0]   # keep only reads where ALL Phred scores are >= 20
```

### Remove reads with Ns or low complexity

```{r}
filter1 <- nFilter(threshold=1)                           # exclude reads with any N
filter2 <- polynFilter(threshold=20, nuc=c("A","T","G","C"))  # exclude low complexity
filter <- compose(filter1, filter2)                       # combine filters
fq[filter(fq)]                                            # apply combined filter
```

---

## Memory-Efficient FASTQ Processing {.scrollable}

For large FASTQ files, streaming avoids loading everything into memory at once.

### Stream or randomly sample reads

```{r}
fq <- yield(FastqStreamer(fastq[1], 50))   # import first 50 reads
fq <- yield(FastqSampler(fastq[1], 50))   # randomly sample 50 reads
```

### Stream through a file with filtering and write output

```{r}
f <- FastqStreamer(fastq[1], 50)    # open streamer with chunk size 50
while (length(fq <- yield(f))) {
    fqsub <- fq[grepl("^TT", sread(fq))]   # keep reads starting with "TT"
    writeFastq(fqsub,
               paste(fastq[1], "sub", sep="_"),
               mode="a",           # append mode — builds file across chunks
               compress=FALSE)
}
close(f)                            # always close the streamer
```

::: {.callout-tip}
**Streaming workflow** is the standard approach for production FASTQ processing:
open a `FastqStreamer`, yield chunks, apply filters/trimming, write output,
repeat until the file is exhausted. This handles files of any size with
constant memory usage.
:::

---

## Range Operations — IRanges and GRanges {.scrollable}

Genomic ranges (chromosome, start, end, strand) are the foundation of most
genome-scale analyses in Bioconductor.

### Three key range containers

| Class | Package | Stores |
|---|---|---|
| `IRanges` | IRanges | integer ranges only |
| `GRanges` | GenomicRanges | ranges + chromosome + strand + metadata |
| `GRangesList` | GenomicRanges | list of GRanges objects |

### Construct a `GRanges` object

```{r}
library(GenomicRanges); library(rtracklayer)

gr <- GRanges(
    seqnames = Rle(c("chr1","chr2","chr1","chr3"), c(1,3,2,4)),
    ranges   = IRanges(1:10, end=7:16, names=head(letters,10)),
    strand   = Rle(strand(c("-","+","*","+","-")), c(1,2,2,3,2)),
    score    = 1:10,
    GC       = seq(1, 0, length=10)
)
```

### Import a GFF/GTF file directly into a `GRanges` object

```{r}
gff <- import.gff("https://cluster.hpcc.ucr.edu/~tgirke/Documents/R_BioCond/Samples/gff3.gff")
seqlengths(gff) <- end(ranges(gff[which(values(gff)[,"type"]=="chromosome"),]))
names(gff) <- 1:length(gff)
gff[1:4,]
seqinfo(gff)

as.data.frame(gff)[1:4, 1:7]    # coerce to data.frame for inspection
```

---

## GRanges — Accessor Functions and Subsetting {.scrollable}

### Accessor functions

```{r}
seqnames(gff)             # chromosome names
ranges(gff)               # IRanges object (start, end, width)
strand(gff)               # strand information
seqlengths(gff)           # chromosome lengths
start(gff[1:4])           # start positions
end(gff[1:4])             # end positions
width(gff[1:4])           # feature widths
```

### Accessing metadata columns

```{r}
values(gff)                                    # all metadata columns
values(gff)[, "type"][1:20]                    # one metadata column
gff[values(gff)[, "type"] == "gene"]          # filter by metadata value
```

### Subsetting and concatenation

```{r}
gff[1:4]                        # subset by index
gff[1:4, c("type", "ID")]       # subset rows and metadata columns
gff[2] <- gff[3]                # replace range
c(gff[1:2], gff[401:402])       # concatenate GRanges objects
```

---

## GRanges — Range Utilities {.scrollable}

### Modify ranges

```{r}
gff <- gff[values(gff)$type != "chromosome"]  # remove chromosome-level entries
strand(gff) <- "*"                             # erase strand information
```

### Set operations on ranges

```{r}
reduce(gff)                                    # collapse overlapping ranges to continuous
gaps(gff)                                      # return uncovered (gap) regions
setdiff(as(seqinfo(gff), "GRanges"), gff)     # gaps — more intuitive approach
disjoin(gff)                                   # return disjoint (non-overlapping) ranges
coverage(gff)                                  # per-base coverage as Rle object
```

### Overlap operations

```{r}
findOverlaps(gff, gff[1:4])                    # index pairs of overlapping ranges
countOverlaps(gff, gff[1:4])[1:40]            # count overlaps per range
subsetByOverlaps(gff, gff[1:4])               # return only overlapping ranges
```

::: {.callout-tip}
`findOverlaps` and `subsetByOverlaps` are among the most frequently used
functions in genomic analyses — they power intersection of features with
peaks, reads, SNPs, and any other range-based data.
:::

### GRangesList — list of GRanges objects

```{r}
sp <- split(gff, seq(along=gff))              # one range per list component
split(gff, seqnames(gff))                     # ranges grouped by chromosome
unlist(sp)                                    # collapse back to GRanges
sp[1:4, "type"]                               # subset GRangesList
lapply(sp[1:4], length)                       # loop over GRangesList like a list
```

---

## Transcript Ranges — TxDb Databases {.scrollable}

`TxDb` (TranscriptDb) objects store transcript, exon, and CDS ranges in a
SQLite database — making annotation retrieval robust and convenient.

### Build a `TxDb` from a GFF file

```{r}
library(txdbmaker)
download.file(
    "https://cluster.hpcc.ucr.edu/~tgirke/Documents/R_BioCond/Samples/gff3.gff",
    "data/gff3.gff"
)
txdb <- makeTxDbFromGFF(
    file       = "data/gff3.gff",
    format     = "gff",
    dataSource = "TAIR",
    organism   = "Arabidopsis thaliana"
)
saveDb(txdb, file="./data/TAIR10.sqlite")    # save for reuse
txdb <- loadDb("./data/TAIR10.sqlite")       # reload later

transcripts(txdb)                            # all transcripts as GRanges
transcriptsBy(txdb, by="gene")               # transcripts grouped by gene
exonsBy(txdb, by="gene")                     # exons grouped by gene
```

### Build a `TxDb` from BioMart

```{r}
library(GenomicFeatures); library(txdbmaker); library(biomaRt)

txdb <- makeTxDbFromBiomart(
    biomart = "plants_mart",
    dataset = "athaliana_eg_gene",
    host    = "https://plants.ensembl.org"
)

# Explore BioMart databases
listMarts(host="plants.ensembl.org")
mymart <- useMart("plants_mart", dataset="athaliana_eg_gene", host="plants.ensembl.org")
listAttributes(mymart)
getBM(attributes=c("ensembl_gene_id", "description"), mart=mymart)[1:4,]
```

---

## Efficient Sequence Parsing from Genome Files {.scrollable}

### `getSeq` — parse sequences for any GRanges annotations

Parse all ranges defined in a `GRanges` object from a genome FASTA file:

```{r}
library(Biostrings); library(Rsamtools)

# Create a random test genome and write to file
gff <- gff[values(gff)$type != "chromosome"]
rand <- DNAStringSet(sapply(
    unique(as.character(seqnames(gff))),
    function(x) paste(sample(c("A","T","G","C"), 200000, replace=TRUE), collapse="")
))
writeXStringSet(DNAStringSet(rand), "./data/test")

# Parse sequences for all annotated ranges
getSeq(FaFile("./data/test"), gff)
```

### `extractTranscriptSeqs` — splice-aware transcript parsing

For multi-exon transcripts, exon sequences must be extracted and concatenated.
`extractTranscriptSeqs` handles this automatically:

```{r}
library(GenomicFeatures); library(Biostrings); library(Rsamtools)

txdb <- loadDb("./data/TAIR10.sqlite")
indexFa("data/test")                                    # index the genome FASTA first

txseq <- extractTranscriptSeqs(
    FaFile("data/test"),
    txdb,
    use.names=TRUE
)
txseq
```

::: {.callout-tip}
Use `getSeq` for simple range extraction (genes, peaks, features).  
Use `extractTranscriptSeqs` when you need spliced transcript sequences
that span multiple exons.
:::

---

## HW05

::: {.callout-important}
**Assignment:** [HW05 — NGS Analysis Basics](https://girke.bioinformatics.ucr.edu/GEN242/assignments/homework/hw06/hw06/)

The homework is based on the sequence analysis, FASTQ processing, and
range operations covered in this tutorial.
:::

---

## Summary — Key Functions by Topic

### Biostrings — sequences

| Function | Purpose |
|---|---|
| `readDNAStringSet()` | import FASTA file |
| `writeXStringSet()` | export to FASTA |
| `reverseComplement()` | reverse complement |
| `translate()` | DNA → protein |
| `matchPattern()` / `vmatchPattern()` | pattern matching (with mismatches) |
| `countPattern()` / `vcountPattern()` | count matches |
| `PWM()` / `matchPWM()` | position weight matrix search |

### ShortRead — FASTQ

| Function | Purpose |
|---|---|
| `readFastq()` | import FASTQ file |
| `sread()`, `quality()`, `id()` | access sequence, quality, ID |
| `seeFastqPlot()` | QC report (systemPipeR) |
| `trimLRPatterns()` | adapter trimming |
| `trimTails()` | quality tail trimming |
| `nFilter()`, `polynFilter()` | N and complexity filters |
| `FastqStreamer()` / `yield()` | memory-efficient streaming |

### GenomicRanges — ranges

| Function | Purpose |
|---|---|
| `import.gff()` | GFF/GTF → GRanges |
| `findOverlaps()` | find overlapping ranges |
| `subsetByOverlaps()` | keep overlapping ranges |
| `reduce()` / `coverage()` | merge / per-base coverage |
| `getSeq()` | extract sequences for ranges |
| `extractTranscriptSeqs()` | spliced transcript sequences |

**Next:** T7 — [Workflow Framework](https://girke.bioinformatics.ucr.edu/GEN242/tutorials/systempiper/systemPipeR_index.html)
