---
title: "HW06 — RNA-Seq Analysis"
sidebar: assignments
---

Source code downloads: [[ .R ]](https://raw.githubusercontent.com/tgirke/GEN242/refs/heads/main/assignments/homework/hw07/hw07.R)

## Overview

This homework covers read counting strategies and differential expression
analysis (DEG) from the
[RNA-Seq Workflow](https://girke.bioinformatics.ucr.edu/GEN242/tutorials/sprnaseq/sprnaseq_index.html)
tutorial. It has three parts:

- **Part A** — Compare unstranded and strand-specific read counting with `featureCounts`
- **Part B** — Count reads over different genomic feature types with `featureCounts`
- **Part C** — DEG analysis with `edgeR` and Venn diagram visualization

All tasks build on the RNA-Seq workflow environment generated with
`systemPipeRdata::genWorkenvir(workflow = "rnaseq")`. Complete the workflow
through the **HISAT2 mapping step** before starting this homework — the sorted
BAM files and `sal` object it produces are required as input throughout.

### Required packages

```{r}
#| eval: false
library(Rsubread)       # featureCounts
library(GenomicFeatures) # makeTxDbFromGFF, intronsByTranscript, fiveUTRsByTranscript
library(systemPipeR)    # run_edgeR, filterDEGs, overLapper, vennPlot
library(edgeR)
```

### Required input objects

Run the following once inside your `rnaseq` workflow directory to prepare
shared inputs used across all tasks:

```{r}
#| eval: false
## Paths to sorted BAM files produced by HISAT2 mapping step
bam_files <- getColumn(sal, step = "hisat2_mapping",
                       "outfiles", column = "samtools_sort_bam")
## Alternative if sal object is not available:
# bam_files <- list.files("results/hisat2_mapping/",
#                         pattern = "sorted.bam$", full.names = TRUE)

## GTF annotation file used in the RNA-Seq workflow
gtf_file <- "data/tair10.gtf"
```

---

## A. Unstranded and Strand-Specific Read Counting

### Background

`featureCounts` from the `Rsubread` package counts reads overlapping
genomic features defined in a GTF file. The `strandSpecific` argument
controls strand awareness:

| Mode | `strandSpecific` | Counts reads on |
|---|---|---|
| Unstranded | `0` | Both strands |
| Sense (positive) strand | `1` | Same strand as feature |
| Antisense (negative) strand | `2` | Opposite strand to feature |

For paired-end RNA-Seq data always set `isPairedEnd = TRUE`. The count
matrix is accessed from the returned list via `$counts`. The `nthreads`
argument controls parallelization — adjust to the number of available cores.

### Task A1 — Generate three count tables

Using `bam_files` as input and `gtf_file` as the annotation, generate three
count tables — one per strand mode. Count at the gene level by grouping exon
features by `gene_id` (`useMetaFeatures = TRUE`).

**1. Unstranded** (`strandSpecific = 0`)

```{r}
#| eval: false
fc_unstranded <- featureCounts(
    files                  = bam_files,
    annot.ext              = gtf_file,
    isGTFAnnotationFile    = TRUE,
    GTF.featureType        = "exon",
    GTF.attrType           = "gene_id",
    useMetaFeatures        = TRUE,
    strandSpecific         = 0,
    isPairedEnd            = TRUE,
    countMultiMappingReads = FALSE,
    nthreads               = 4
)
unstranded <- fc_unstranded$counts
unstranded[1:4, ]
```

**2. Sense (positive) strand** (`strandSpecific = 1`)

```{r}
#| eval: false
fc_stranded_pos <- featureCounts(
    files                  = bam_files,
    annot.ext              = gtf_file,
    isGTFAnnotationFile    = TRUE,
    GTF.featureType        = "exon",
    GTF.attrType           = "gene_id",
    useMetaFeatures        = TRUE,
    strandSpecific         = 1,
    isPairedEnd            = TRUE,
    countMultiMappingReads = FALSE,
    nthreads               = 4
)
stranded_pos <- fc_stranded_pos$counts
stranded_pos[1:4, ]
```

**3. Antisense (negative) strand** (`strandSpecific = 2`)

```{r}
#| eval: false
fc_stranded_neg <- featureCounts(
    files                  = bam_files,
    annot.ext              = gtf_file,
    isGTFAnnotationFile    = TRUE,
    GTF.featureType        = "exon",
    GTF.attrType           = "gene_id",
    useMetaFeatures        = TRUE,
    strandSpecific         = 2,
    isPairedEnd            = TRUE,
    countMultiMappingReads = FALSE,
    nthreads               = 4
)
stranded_neg <- fc_stranded_neg$counts
stranded_neg[1:4, ]
```

Store the count matrices in variables named exactly **`unstranded`**,
**`stranded_pos`**, and **`stranded_neg`**.

### Task A2 — Compare strand-specific counts to unstranded

Test whether the two strand-specific count tables sum to approximately the
same values as the unstranded count table. Calculate for each
strand-specific result the mean mapping frequency as a percentage of the
unstranded counts:

```{r}
#| eval: false
perc_pos <- mean((stranded_pos / unstranded) * 100, na.rm = TRUE)
perc_neg <- mean((stranded_neg / unstranded) * 100, na.rm = TRUE)
perc_pos
perc_neg
```

Store results in variables named exactly **`perc_pos`** and **`perc_neg`**.

**Interpretation:** if both values are close to 50%, the RNA-Seq library is
very likely **unstranded**. If one value dominates (e.g. ~100% sense,
~0% antisense), the library is strand-specific. For paired-end data expect
very similar but not necessarily identical results between the sum of stranded
counts and the unstranded total.

### Task A3 — Biological interpretation of strand-specific counting

Include comment text in your script addressing the following four points:

1. When should strand-specific counting be used instead of unstranded counting?
2. How does strand-specific counting help resolve reads from overlapping genes
   encoded on opposite strands?
3. What biological information can antisense strand counting reveal?
4. How can comparing stranded and unstranded results determine whether an
   RNA-Seq experiment used a strand-specific library preparation protocol?

---

## B. Read Counting for Different Feature Types

### Background

By changing `GTF.featureType` and `useMetaFeatures` in `featureCounts`,
reads can be counted over different genomic features (genes, exons, introns,
UTRs). Importantly, `featureCounts` accepts `GRanges` and `GRangesList`
objects directly via the `annot.ext` argument — no conversion to a data
frame is needed. This makes it straightforward to pass intron or UTR ranges
derived from a `TxDb` object directly to `featureCounts`. Set
`isGTFAnnotationFile = FALSE` whenever `annot.ext` is a range object rather
than a GTF file path.

### Task B — Count reads over five feature types

Compute **sense strand** (`strandSpecific = 1`) count tables for the
following five feature types. Use `isPairedEnd = TRUE` for genes and exon
features. For introns and UTRs use `isPairedEnd = FALSE` (single-end mode
since these are sub-gene features).

**1. Genes** — exons grouped by gene (`useMetaFeatures = TRUE`)

```{r}
#| eval: false
fc_genes <- featureCounts(
    files                  = bam_files,
    annot.ext              = gtf_file,
    isGTFAnnotationFile    = TRUE,
    GTF.featureType        = "exon",
    GTF.attrType           = "gene_id",
    useMetaFeatures        = TRUE,
    strandSpecific         = 1,
    isPairedEnd            = TRUE,
    countMultiMappingReads = FALSE,
    nthreads               = 4
)
gene_ranges_countDF <- fc_genes$counts
gene_ranges_countDF[1:4, ]
```

**2. Exons** — individual exon level (`useMetaFeatures = FALSE`)

```{r}
#| eval: false
fc_exons <- featureCounts(
    files                  = bam_files,
    annot.ext              = gtf_file,
    isGTFAnnotationFile    = TRUE,
    GTF.featureType        = "exon",
    GTF.attrType           = "gene_id",
    useMetaFeatures        = FALSE,
    strandSpecific         = 1,
    isPairedEnd            = TRUE,
    countMultiMappingReads = FALSE,
    nthreads               = 4
)
exon_ranges_countDF <- fc_exons$counts
exon_ranges_countDF[1:4, ]
```

**3. Exons by genes** — same as Task B1; explicitly assigned for direct
comparison with the `exonsBy(txdb, by="gene")` approach used in tutorials:

```{r}
#| eval: false
exonByg_ranges_countDF <- fc_genes$counts   # reuses result from Task B1
exonByg_ranges_countDF[1:4, ]
```

**4. Introns by transcripts** — derive intron ranges from `txdb` and pass
the `GRangesList` directly to `featureCounts` via `annot.ext`

```{r}
#| eval: false
txdb          <- makeTxDbFromGFF(gtf_file, format = "gtf")
intron_ranges <- intronsByTranscript(txdb, use.names = TRUE)

## GRangesList can be passed directly — no data frame conversion needed
fc_introns <- featureCounts(
    files                  = bam_files,
    annot.ext              = intron_ranges,
    isGTFAnnotationFile    = FALSE,
    strandSpecific         = 1,
    isPairedEnd            = FALSE,
    countMultiMappingReads = FALSE,
    nthreads               = 4
)
intron_ranges_countDF <- fc_introns$counts
intron_ranges_countDF[1:4, ]
```

**5. 5'-UTRs by transcripts** — same approach as introns, passing the
`GRangesList` directly

```{r}
#| eval: false
fiveUTR_ranges <- fiveUTRsByTranscript(txdb, use.names = TRUE)

fc_fiveUTR <- featureCounts(
    files                  = bam_files,
    annot.ext              = fiveUTR_ranges,
    isGTFAnnotationFile    = FALSE,
    strandSpecific         = 1,
    isPairedEnd            = FALSE,
    countMultiMappingReads = FALSE,
    nthreads               = 4
)
fiveUTR_ranges_countDF <- fc_fiveUTR$counts
fiveUTR_ranges_countDF[1:4, ]
```

Use the **exact variable names** shown above:
`gene_ranges_countDF`, `exon_ranges_countDF`, `exonByg_ranges_countDF`,
`intron_ranges_countDF`, `fiveUTR_ranges_countDF`.

---

## C. DEG Analysis with edgeR

### Background

`run_edgeR()` from `systemPipeR` wraps the standard edgeR DEG workflow.
Here you will run it on two count tables (unstranded vs sense strand) and
compare the resulting DEG sets visually using 4-way Venn diagrams.

### Task C — DEG analysis and Venn diagram comparison

**1. Run edgeR on both count tables**

```{r}
#| eval: false
cmp <- readComp(stepsWF(sal)[["hisat2_mapping"]],
                format = "matrix", delim = "-")
## Alternative if sal is not available:
# cmp <- readComp(file = "targetsPE.txt", format = "matrix", delim = "-")

## DEG analysis with unstranded count table
edgeDF_unstranded <- run_edgeR(
    countDF     = unstranded,
    targets     = targetsWF(sal)[["hisat2_mapping"]],
    cmp         = cmp[[1]],
    independent = FALSE,
    mdsplot     = ""
)

## DEG analysis with sense strand count table
edgeDF_pos <- run_edgeR(
    countDF     = stranded_pos,
    targets     = targetsWF(sal)[["hisat2_mapping"]],
    cmp         = cmp[[1]],
    independent = FALSE,
    mdsplot     = ""
)
```

**2. Generate and save 4-way Venn diagrams**

```{r}
#| eval: false
## (1) Venn diagram — unstranded count table
DEG_list_unstranded <- filterDEGs(
    degDF  = edgeDF_unstranded,
    filter = c(Fold = 2, FDR = 40),
    plot   = FALSE
)
vennsetup   <- overLapper(DEG_list_unstranded$Up[6:9],   type = "vennsets")
vennsetdown <- overLapper(DEG_list_unstranded$Down[6:9], type = "vennsets")
pdf("results/DEGcount_unstranded.pdf")
vennPlot(list(vennsetup, vennsetdown),
         mymain = "", mysub = "", colmode = 2, ccol = c("blue", "red"))
dev.off()

## (2) Venn diagram — sense strand count table
DEG_list_pos <- filterDEGs(
    degDF  = edgeDF_pos,
    filter = c(Fold = 2, FDR = 40),
    plot   = FALSE
)
vennsetup   <- overLapper(DEG_list_pos$Up[6:9],   type = "vennsets")
vennsetdown <- overLapper(DEG_list_pos$Down[6:9], type = "vennsets")
pdf("results/DEGcount_pos.pdf")
vennPlot(list(vennsetup, vennsetdown),
         mymain = "", mysub = "", colmode = 2, ccol = c("blue", "red"))
dev.off()
```

Both PDF files must be committed to the repository at:

- `results/DEGcount_unstranded.pdf`
- `results/DEGcount_pos.pdf`

---

## Homework Submission

Submit **one R script** named **`HW6.R`** and the **two Venn diagram PDFs**
to your private GitHub homework repository at the following paths:

```
Homework/HW6/HW6.R
results/DEGcount_unstranded.pdf
results/DEGcount_pos.pdf
```

### Requirements for full credit

1. Use **`featureCounts()`** from `Rsubread` for all read counting (Parts A and B)
2. Define all variables with the **exact names** specified throughout
3. Include **comment text** for all four points of Task A3
4. Commit both **Venn diagram PDF files** alongside `HW7.R`
5. Script must be well structured following the recommended layout below

### Recommended script structure

```{r}
#| eval: false
## ─────────────────────────────────────────────────────────────
## HW7: RNA-Seq Analysis
## GEN242, Spring 2026
## Author: <your name>
## ─────────────────────────────────────────────────────────────

library(Rsubread); library(GenomicFeatures)
library(systemPipeR); library(edgeR)

## Input
bam_files <- getColumn(sal, step = "hisat2_mapping",
                       "outfiles", column = "samtools_sort_bam")
gtf_file  <- "data/tair10.gtf"

## ── Part A: Strand-specific read counting ─────────────────────

## Task A1 — unstranded (strandSpecific = 0)
fc_unstranded <- featureCounts(files = bam_files, annot.ext = gtf_file,
                               isGTFAnnotationFile = TRUE,
                               GTF.featureType = "exon", GTF.attrType = "gene_id",
                               useMetaFeatures = TRUE, strandSpecific = 0,
                               isPairedEnd = TRUE, nthreads = 4)
unstranded <- fc_unstranded$counts

## Task A1 — sense strand (strandSpecific = 1)
fc_stranded_pos <- featureCounts( <fill in arguments, strandSpecific = 1> )
stranded_pos <- fc_stranded_pos$counts

## Task A1 — antisense strand (strandSpecific = 2)
fc_stranded_neg <- featureCounts( <fill in arguments, strandSpecific = 2> )
stranded_neg <- fc_stranded_neg$counts

## Task A2 — percentage comparison
perc_pos <- mean((stranded_pos / unstranded) * 100, na.rm = TRUE)
perc_neg <- mean((stranded_neg / unstranded) * 100, na.rm = TRUE)

## Task A3 — biological interpretation (replace ... with your answers)
## 1. Strand-specific counting should be used when the library is stranded because ...
## 2. For overlapping genes on opposite strands, strand-specific counting allows ...
## 3. Antisense strand counting can reveal biological events such as ...
## 4. Comparing stranded and unstranded results determines strandedness because ...

## ── Part B: Feature type counting ────────────────────────────

## B1: Genes
gene_ranges_countDF    <- featureCounts( <fill in, useMetaFeatures = TRUE> )$counts

## B2: Exons (individual)
exon_ranges_countDF    <- featureCounts( <fill in, useMetaFeatures = FALSE> )$counts

## B3: Exons by genes
exonByg_ranges_countDF <- gene_ranges_countDF

## B4: Introns
txdb          <- makeTxDbFromGFF(gtf_file, format = "gtf")
intron_ranges <- intronsByTranscript(txdb, use.names = TRUE)
intron_ranges_countDF  <- featureCounts( <fill in, annot.ext = intron_ranges> )$counts

## B5: 5'-UTRs
fiveUTR_ranges <- fiveUTRsByTranscript(txdb, use.names = TRUE)
fiveUTR_ranges_countDF <- featureCounts( <fill in, annot.ext = fiveUTR_ranges> )$counts

## ── Part C: DEG analysis ──────────────────────────────────────

cmp <- readComp(stepsWF(sal)[["hisat2_mapping"]], format = "matrix", delim = "-")

edgeDF_unstranded <- run_edgeR(countDF = unstranded,
                               targets = targetsWF(sal)[["hisat2_mapping"]],
                               cmp = cmp[[1]], independent = FALSE, mdsplot = "")
edgeDF_pos        <- run_edgeR(countDF = stranded_pos,
                               targets = targetsWF(sal)[["hisat2_mapping"]],
                               cmp = cmp[[1]], independent = FALSE, mdsplot = "")

## Venn diagrams → results/DEGcount_unstranded.pdf and results/DEGcount_pos.pdf
## <add filterDEGs, overLapper, vennPlot calls here>
```

---

## Due Date

This homework is due **Thur, May 14th at 6:00 PM**.

## Homework Solutions

To be posted after the due date.
