---
title: "Introduction to R (T3)"
subtitle: "R basics — syntax, data types, and control structures"
author: "Thomas Girke"
date: last-modified
sidebar: tutorials       # activates the Tutorials sidebar defined in _quarto.yml
---

## Overview

This tutorial introduces the basic building blocks of R programming.
Code chunks below execute at build time; outputs — plots, tables, printed
results — are embedded directly in the page.

## Vectors and data types

```{r}
#| label: basic-vectors

x       <- c(1, 2, 3, 4, 5)
species <- c("Arabidopsis", "Human", "Yeast", "Mouse")
gc      <- c(AT = 0.40, GC = 0.60)
gc
```

## Interactive tables with DT

```{r}
#| label: interactive-table
library(dplyr)
library(DT)

set.seed(42)
gene_df <- data.frame(
  GeneID = paste0("AT", sample(1000:9999, 20)),
  logFC  = round(rnorm(20, sd = 2), 3),
  pvalue = round(runif(20, 0, 0.05), 4),
  FDR    = round(runif(20, 0, 0.1),  4)
)

datatable(gene_df, filter = "top",
          options = list(pageLength = 5, scrollX = TRUE),
          caption = "Table 1: Simulated DEG results")
```

## Figures

```{r}
#| label: fig-volcano
#| fig-cap: "Volcano plot of simulated results"
library(ggplot2)

gene_df |>
  mutate(sig = FDR < 0.05 & abs(logFC) > 1) |>
  ggplot(aes(logFC, -log10(pvalue), color = sig)) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_color_manual(values = c("grey60", "#e63946"), labels = c("NS", "Sig")) +
  labs(x = "log2 Fold Change", y = "-log10(p-value)", color = NULL) +
  theme_minimal(base_size = 13)
```

## Python interoperability

::: {.callout-tip}
## Freezing expensive computations
Add `#| cache: true` to chunks that call slow Bioconductor functions.
Results are reused on subsequent builds unless the source changes.
:::

## Session info

```{r}
#| label: session-info
#| code-fold: true
#| code-summary: "Show session info"
sessionInfo()
```
