---
title: "HW05 — NGS Analysis Basics"
sidebar: assignments
---

Source code downloads: [[ .R ]](https://raw.githubusercontent.com/tgirke/GEN242/refs/heads/main/assignments/homework/hw05/hw05.R)

## Overview

This homework covers two topics from the 
[NGS Analysis Basics](https://girke.bioinformatics.ucr.edu/GEN242/tutorials/rsequences/rsequences_index.html) 
tutorial: **(A)** extending a demultiplexing function to also trim low-quality 
read tails, and **(B)** parsing genomic sequences using GFF annotation and 
Bioconductor range operations. Both parts require the `Biostrings`, 
`ShortRead`, `GenomicRanges`, and `rtracklayer` Bioconductor packages.

Install required packages if needed:

```{r}
#| eval: false
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(c("Biostrings", "ShortRead", "GenomicRanges", "rtracklayer"))
```

---

## A. Demultiplexing with Quality Trimming

### Background

Demultiplexing splits a multiplexed FASTQ file into separate files based on 
barcode sequences at the start of each read. A common additional step is to 
remove low-quality bases from the 3' end of each read (quality tail trimming) 
before writing the output.

The following starter function handles **demultiplexing only**. Your task is 
to extend it to also perform **quality tail trimming**.

### Starter function (demultiplexing only)

```{r}
#| eval: false
library(ShortRead)

demultiplex <- function(x, barcode, nreads) {
    f <- FastqStreamer(x, nreads)
    while(length(fq <- yield(f))) {
        for(i in barcode) {
            pattern <- paste("^", i, sep = "")
            fqsub <- fq[grepl(pattern, sread(fq))]
            if(length(fqsub) > 0) {
                writeFastq(fqsub, paste(x, i, sep = "_"),
                           mode = "a", compress = FALSE)
            }
        }
    }
    close(f)
}
```

### Task A — Extend `demultiplex` to trim low-quality tails

Write a new function named **`demultiplex_trim`** that extends the starter 
function above with a quality trimming step. The function must have the 
following **exact signature**:

```{r}
#| eval: false
demultiplex_trim <- function(x, barcode, nreads, cutoff = 20) { ... }
```

**Arguments:**

| Argument | Type | Description |
|---|---|---|
| `x` | character | Path to input FASTQ file |
| `barcode` | character vector | One or more barcode sequences to match at read start |
| `nreads` | integer | Number of reads to process per chunk (passed to `FastqStreamer`) |
| `cutoff` | integer | Minimum quality score threshold for trimming (default: 20) |

**Required behavior:**

1. Stream the FASTQ file in chunks of `nreads` reads
2. For each barcode, subset reads whose sequence starts with that barcode (same as starter)
3. **Trim low-quality bases from the 3' tail** of each matched read before writing — remove trailing bases whose quality score falls below `cutoff`
4. Write trimmed, barcode-matched reads to a file named `<x>_<barcode>` using `writeFastq(..., mode="a", compress=FALSE)`

**Useful functions for quality trimming:**

```{r}
#| eval: false
## trimTails() from ShortRead trims bases from the 3' end
## where quality drops below a threshold
## Arguments: object, k (min consecutive good bases), a (quality character threshold)
trimTails(fqsub, k = 2, a = encoding(cutoff))

## Alternatively, use trimEnds() for a simpler threshold approach
## PhredQuality encoding: quality score 20 = ASCII '5'
## Use encoding() to convert numeric cutoff to the correct character
quality(fqsub)  # inspect quality scores
encoding(quality(fqsub))  # see score-to-character mapping
```

**Test data:** Download the FASTQ test files from the tutorial:

```{r}
#| eval: false
## Download FASTQ test files used in the NGS tutorial
library(ShortRead)
fastq <- dir(system.file("extdata", package = "ShortRead"),
             pattern = "fastq$", full.names = TRUE)

## Run your function on the first test file
demultiplex_trim(x = fastq[1], barcode = c("TT", "AA", "GG"),
                 nreads = 50, cutoff = 20)
```

**Expected output:** One FASTQ file per barcode (e.g. `<fastq[1]>_TT`, 
`<fastq[1]>_AA`, `<fastq[1]>_GG`) in the working directory, containing 
trimmed reads that matched the respective barcode.

---

## B. Sequence Parsing from GFF and Genome

### Background

Genomic annotation files in GFF format define the coordinates of genes and 
other features on a reference genome. In this task you will use Bioconductor 
range operations to extract gene sequences, reduce overlapping ranges, and 
generate intergenic sequences from the *Halobacterium sp.* genome.

### Download data

Download the GFF annotation file and genome FASTA for *Halobacterium sp.*:

```{r}
#| eval: false
library(Biostrings)
library(GenomicRanges)
library(rtracklayer)

dir.create("data", showWarnings = FALSE)

download.file(
    "https://ftp.ncbi.nlm.nih.gov/genomes/archive/old_genbank/Bacteria/Halobacterium_sp_uid217/AE004437.gff",
    "data/AE004437.gff"
)
download.file(
    "https://ftp.ncbi.nlm.nih.gov/genomes/archive/old_genbank/Bacteria/Halobacterium_sp_uid217/AE004437.fna",
    "data/AE004437.fna"
)

## Import genome and GFF
chr    <- readDNAStringSet("data/AE004437.fna")
gff    <- import("data/AE004437.gff")
```

### Task B1 — Extract gene sequences and translate to protein

Extract all gene ranges from the GFF, parse their sequences from the genome, 
and translate them into protein sequences. Account for strand: genes on the 
`+` strand are translated directly; genes on the `-` strand must be 
reverse-complemented before translation.

Write the combined protein sequences to a file named **`data/mypep.fasta`**.

**Required variable names** (used by the grading script):

| Variable | Content |
|---|---|
| `gffgene` | GRanges object — gene features only (type == "gene") |
| `gene` | `DNAStringSet` of gene sequences parsed from genome |
| `p1` | `AAStringSet` of translated proteins from `+` strand genes |
| `p2` | `AAStringSet` of translated proteins from `-` strand genes |

**Useful commands:**

```{r}
#| eval: false
## Filter GFF to gene features only
gffgene <- gff[values(gff)[, "type"] == "gene"]

## Extract gene sequences using Views
gene <- DNAStringSet(Views(chr[[1]],
                           IRanges(start(gffgene), end(gffgene))))
names(gene) <- values(gffgene)[, "locus_tag"]

## Translate + strand genes
pos <- values(gffgene[strand(gffgene) == "+"])[, "locus_tag"]
p1  <- translate(gene[names(gene) %in% pos])
names(p1) <- names(gene[names(gene) %in% pos])

## Translate - strand genes (reverse complement first)
neg <- values(gffgene[strand(gffgene) == "-"])[, "locus_tag"]
p2  <- translate(reverseComplement(gene[names(gene) %in% neg]))
names(p2) <- names(gene[names(gene) %in% neg])

## Write all proteins to FASTA
writeXStringSet(c(p1, p2), "./data/mypep.fasta")
```

### Task B2 — Reduce overlapping gene ranges

Collapse overlapping gene ranges using the `reduce()` function and parse 
the resulting non-redundant sequences from the genome.

**Required variable names:**

| Variable | Content |
|---|---|
| `gffgene_reduced` | `GRanges` result of `reduce(gffgene)` |
| `gene_reduced` | `DNAStringSet` of sequences for reduced ranges |

**Useful commands:**

```{r}
#| eval: false
## Reduce overlapping ranges
gffgene_reduced <- reduce(gffgene)

## Parse sequences for reduced ranges
gene_reduced <- DNAStringSet(Views(chr[[1]],
                    IRanges(start(gffgene_reduced),
                            end(gffgene_reduced))))
```

### Task B3 — Generate intergenic ranges and parse sequences

Compute the intergenic regions (gaps between genes) and parse their sequences 
from the genome.

**Required variable names:**

| Variable | Content |
|---|---|
| `intergenic` | `GRanges` or `IRanges` of intergenic regions |
| `gene_intergenic` | `DNAStringSet` of intergenic sequences |

**Useful commands:**

```{r}
#| eval: false
## Compute gaps between reduced gene ranges
intergenic <- gaps(gffgene_reduced)

## Parse intergenic sequences
gene_intergenic <- DNAStringSet(Views(chr[[1]],
                       IRanges(start(intergenic),
                               end(intergenic))))
```

---

## Homework Submission

Submit **one R script** named **`HW5.R`** to your private GitHub homework 
repository at the following exact path:

```
Homework/HW5/HW5.R
```

### Requirements for full credit

Your script must:

1. Define the function **`demultiplex_trim(x, barcode, nreads, cutoff = 20)`** 
   with the quality trimming step implemented
2. Include a **usage example** showing how to call `demultiplex_trim` on the 
   ShortRead test FASTQ files
3. Implement all three **Task B** steps using the required variable names 
   (`gffgene`, `gene`, `p1`, `p2`, `gffgene_reduced`, `gene_reduced`, 
   `intergenic`, `gene_intergenic`)
4. Write the protein FASTA output to **`data/mypep.fasta`**
5. Include **brief comments** explaining each step

### Recommended script structure

```{r}
#| eval: false
## ─────────────────────────────────────────────────────────────
## HW5: NGS Analysis Basics
## GEN242, Spring 2026
## Author: <your name>
## ─────────────────────────────────────────────────────────────

library(ShortRead)
library(Biostrings)
library(GenomicRanges)
library(rtracklayer)

## ── Part A: Demultiplexing with quality trimming ──────────────

demultiplex_trim <- function(x, barcode, nreads, cutoff = 20) {
    ## ... your implementation ...
}

## Usage example
# fastq <- dir(system.file("extdata", package = "ShortRead"),
#              pattern = "fastq$", full.names = TRUE)
# demultiplex_trim(x = fastq[1], barcode = c("TT", "AA", "GG"),
#                 nreads = 50, cutoff = 20)

## ── Part B: Sequence parsing from GFF and genome ─────────────

dir.create("data", showWarnings = FALSE)

## Download data
# download.file("https://ftp.ncbi.nlm.nih.gov/genomes/archive/old_genbank/Bacteria/Halobacterium_sp_uid217/AE004437.gff", "data/AE004437.gff")
# download.file("https://ftp.ncbi.nlm.nih.gov/genomes/archive/old_genbank/Bacteria/Halobacterium_sp_uid217/AE004437.fna", "data/AE004437.fna")

## Task B1: Extract gene sequences and translate to protein
## ... your implementation using gffgene, gene, p1, p2 ...

## Task B2: Reduce overlapping gene ranges
## ... your implementation using gffgene_reduced, gene_reduced ...

## Task B3: Generate intergenic ranges
## ... your implementation using intergenic, gene_intergenic ...
```

---

## Due Date

This homework is due **Tuesday, May 6th at 6:00 PM**.

## Homework Solutions

To be posted after the due date.
