---
title: "HW04: Programming in R"
sidebar: assignments
---

## Overview

This homework involves writing three R functions for transforming and 
translating DNA sequences. All functions should be implemented in a single 
well-structured and annotated R script submitted to your private GitHub 
repository. Read the instructions for each task carefully. The expected 
function names, arguments, and output formats are detailed below and 
will be used for grading.

---

## A. Reverse and Complement of DNA

### Task 1 — Write the `RevComp` function

Write a function named exactly **`RevComp`** that accepts a single DNA 
sequence string and returns a tranformed version of it. The function must 
include an argument named **`type`** that controls which transformation 
is applied:

| `type` value | Transformation |
|---|---|
| `"rev"` | Return only the **reversed** sequence |
| `"comp"` | Return only the **complemented** sequence |
| `"revcomp"` | Return the **reverse complement** (reversed then complemented) |

The default value of `type` should be `"revcomp"`.

**Required function signature:**
```r
RevComp <- function(x, type = "revcomp") { ... }
```

**Expected results for test sequence `"ATGCATTGGACGTTAG"`:**

```r
x <- "ATGCATTGGACGTTAG"
RevComp(x, type = "rev")      # → "GATTGCAGGTTACGTA"
RevComp(x, type = "comp")     # → "TACGTAACCTGCAATC"
RevComp(x, type = "revcomp")  # → "CTAACGTCCAATGCAT"
RevComp(x)                    # → "CTAACGTCCAATGCAT"  (default)
```

**Useful R building blocks for your implementation:**

```{r}
#| eval: false
x <- "ATGCATTGGACGTTAG"

## Step 1: vectorize the string into individual characters
x <- substring(x, 1:nchar(x), 1:nchar(x))
x
## [1] "A" "T" "G" "C" "A" "T" "T" "G" "G" "A" "C" "G" "T" "T" "A" "G"

## Step 2: reverse the character vector
x <- rev(x)
x
## [1] "G" "A" "T" "T" "G" "C" "A" "G" "G" "T" "T" "A" "C" "G" "T" "A"

## Step 3: collapse back into a single string
x <- paste(x, collapse = "")
x
## [1] "GATTGCAGGTTACGTA"

## Step 4: complement using base substitution (A↔T, G↔C)
chartr("ATGC", "TACG", x)
## [1] "CTAACGTCCAATGCAT"
```

---

### Task 2 — Vectorize `RevComp` and export to FASTA

Write two additional functions:

**(a)** A function named **`RevCompVector`** that applies `RevComp` to a 
**vector of DNA sequences** (multiple sequences at once). It should accept 
the same `type` argument and pass it through to `RevComp`.

**Required function signature:**
```r
RevCompVector <- function(x, type = "revcomp") { ... }
```

**Expected behavior:**
```r
seqs <- c("ATGCATTGGACGTTAG", "TTGGCAATCGA", "GCTAGCTA")
RevCompVector(seqs, type = "rev")
## [1] "GATTGCAGGTTACGTA" "AGCTAACGGTT"      "ATCGATCG"
```

**(b)** A function named **`WriteFasta`** that saves a named vector of DNA 
sequences to a file in standard FASTA format. Each sequence should be 
preceded by a header line starting with `>` followed by the sequence name.

**Required function signature:**
```r
WriteFasta <- function(seqs, file) { ... }
```

**Expected output format** in the saved file:
```
>seq1
ATGCATTGGACGTTAG
>seq2
TTGGCAATCGA
>seq3
GCTAGCTA
```

**Example usage:**
```r
seqs <- c(seq1 = "ATGCATTGGACGTTAG",
          seq2 = "TTGGCAATCGA",
          seq3 = "GCTAGCTA")
WriteFasta(seqs, file = "myseqs.fasta")
```

If the input vector has no names, the function should assign default names 
`seq1`, `seq2`, ... automatically.

---

## B. Translate DNA into Protein

### Task 3 — Write a DNA translation function

Write a function named **`TranslateDNA`** that translates one or more DNA 
sequences into protein sequences using the standard genetic code. The 
function should translate in all **three reading frames** (frame 1, 2, 
and 3) and return all translations.

**Required function signature:**
```r
TranslateDNA <- function(x) { ... }
```

Where `x` can be a single sequence string or a vector of sequences.

**Return value:** a named list where each element contains 
the three reading frame translations for one input sequence, labeled 
`frame1`, `frame2`, `frame3`. Stop codons should be represented as `*`.

**Expected output for a single sequence:**
```r
TranslateDNA("ATGCATTGGACGTTAG")
## $frame1
## [1] "MHWT*"
## $frame2
## [1] "CIGR"  (or similar, depending on frame shift)
## $frame3
## [1] "ALDS"  (or similar, depending on frame shift)
```

**Useful R building blocks for your implementation:**

```{r}
#| eval: false
## Import the genetic code lookup table
AAdf <- read.table(
    file   = "http://faculty.ucr.edu/~tgirke/Documents/R_BioCond/My_R_Scripts/AA.txt",
    header = TRUE, sep = "\t"
)
AAdf[1:4, ]
##   Codon AA_1 AA_3 AA_Full AntiCodon
## 1   TCA    S  Ser  Serine       TGA
## 2   TCG    S  Ser  Serine       CGA
## 3   TCC    S  Ser  Serine       GGA
## 4   TCT    S  Ser  Serine       AGA

## Create named vector: codon → single-letter amino acid
AAv <- as.character(AAdf[, 2])
names(AAv) <- AAdf[, 1]

## Tripletize and translate (for a sequence x already split into characters)
y <- gsub("(...)", "\\1_", x)   # insert _ after every 3 chars
y <- unlist(strsplit(y, "_"))    # split on _
y <- y[grep("^...$", y)]        # keep only complete triplets
AAv[y]                           # look up amino acids by codon name
## GAT TGC AGG TTA CGT
## "D" "C" "R" "L" "R"
```

To translate in **frame 2**, skip the first character of the sequence 
before tripletizing. To translate in **frame 3**, skip the first two 
characters. Use `substring(seq, start)` to shift the reading frame.

---

## Homework Submission

Submit **one R script** named **`HW4.R`** to your 
private GitHub homework repository at the following exact path:

```
Homework/HW4/HW4.R
```

Optionally, students who wish to demonstrate their functions in a rendered document can additionally submit a HW4.qmd file in the same directory. The .qmd file should source the HW4.R script and execute the functions from it, for example:

```{r}
#| eval: false
source("HW4.R")

## Run RevComp
RevComp("ATGCATTGGACGTTAG", type = "revcomp")

## Run TranslateDNA
TranslateDNA("ATGCATTGGACGTTAG")
```
Note that the `.qmd` file is an optional add-on and will not be used for grading. Grading is performed exclusively on `HW4.R`. The `HW4.R` file must still be fully self-contained and structured as specified above regardless of whether a `.qmd` file is also submitted.

### Requirements for full credit

Your submitted script must:

1. Define all four functions with the **exact names** specified:
   - `RevComp(x, type = "revcomp")`
   - `RevCompVector(x, type = "revcomp")`
   - `WriteFasta(seqs, file)`
   - `TranslateDNA(x)`

2. Include a **usage example** for each function as commented-out code in a `## Usage` section at the end of each task.

3. Include **brief comments** explaining what each function does and what 
   its arguments mean

4. Be **runnable without errors** — source the script with `source("HW4.R")` 
   before submitting to check this

5. Handle the case where the input sequence vector has **no names** 
   (auto-assign `seq1`, `seq2`, ... in `WriteFasta`)

### Recommended script structure

```r
## ─────────────────────────────────────────────────────────────
## HW4: DNA Sequence Manipulation and Translation
## GEN242, Spring 2026
## Author: <your name>
## Date: <submission date>
## ─────────────────────────────────────────────────────────────

## Task 1: RevComp function
RevComp <- function(x, type = "revcomp") {
    ## ... your implementation ...
}

## Task 2a: RevCompVector function
RevCompVector <- function(x, type = "revcomp") {
    ## ... your implementation ...
}

## Task 2b: WriteFasta function
WriteFasta <- function(seqs, file) {
    ## ... your implementation ...
}

## Task 3: TranslateDNA function
TranslateDNA <- function(x) {
    ## ... your implementation ...
}

## ── Usage examples ────────────────────────────────────────────

## Task 1
# RevComp("ATGCATTGGACGTTAG", type = "rev")
# RevComp("ATGCATTGGACGTTAG", type = "comp")
# RevComp("ATGCATTGGACGTTAG", type = "revcomp")

## Task 2
# seqs <- c(seq1 = "ATGCATTGGACGTTAG", seq2 = "TTGGCAATCGA")
# RevCompVector(seqs, type = "revcomp")
# WriteFasta(seqs, file = "myseqs.fasta")

## Task 3
# TranslateDNA("ATGCATTGGACGTTAG")
# TranslateDNA(c("ATGCATTGGACGTTAG", "TTGGCAATCGA"))
```

---

## Due Date

This homework is due **Tuesday, April 28th at 6:00 PM**.

## Homework Solutions

To be posted after the due date.
